---
title: "Introduction to R"
date: "`r Sys.Date()`"
output: html_document
---

### Using R as a calculator

The simplest thing you could do with R is to do arithmetic. 

__Practice__: Enter each of the following commands and confirm that the response
is the correct answer.  

```{r}  
1 + 2 
16*9
sqrt(2)
20/5
18.5 - 7.21
3 %% 2 ## what is this doing?
```


### Creating objects

These operations, however, are not very interesting. To do more useful things in
R, we need to assign values to an object. To create an object, we tell R the
object's name, followed by an assignment arrow (`<-`), and finally the value of 
the object. This would look something like this:

`x <- 6`

Once we execute/run this line of code, we notice that a new object appears in 
our environment window. This window shows all of the objects that you have 
created during your R session. The value of `x` appears next to it, since it 
is a scalar. 


## A note on objects vs. variables

> What are known as `objects` in R are known as `variables` in many other
> programming languages. Depending on the context, `object` and `variable` can
> have drastically different meanings. However, in this lesson, the two words
> are used synonymously. For more information see:
> [https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects](https://cran.r-project.org/doc/manuals/r-release/R-lang.html#Objects)

When assigning a value to an object, R does not print anything. You
can force R to print the value by using parentheses or by typing
the object name:

```{r}
year <- 2016    # doesn't print anything
(year <- 2016)  # putting parenthesis around the call prints the value of year
year         # and so does typing the name of the object
```

Now that R has `year` in memory, we can do arithmetic with it. For
instance, we may want to find the number of years after 1989 it is:

```{r}
year - 1989
```

We can also change an object's value by assigning it a new one:

```{r}
year <- 2005
year - 1989
```

This means that assigning a value to one object does not change the values of
other objects. For example, let's store the plot's area in acres
in a new object, `years_post_birthday`:

```{r}
years_post_birthday <- year - 1989
```

and then change `year` to 2002.

```{r}
year <- 2002
```

### Your turn!

### Exercise

What do you think is the current content of the object `years_post_birthday`?
16 or 13? 


### Adding code comments

All programming languages allow the programmer to include comments in their
code. To do this in R we use the `#` character. Anything to the right of the `#`
sign and up to the end of the line is treated as a comment and is ignored by R.
You can start lines with comments or include them after any code on the line.

```{r}
ms_year <- 2016    # Year got M.S.
phd_year <- ms_year + 4 # Year got Ph.D. 
```


### Your turn! 

### Exercise

Create two variables `length` and `width` and assign them values. It should be
noted that, because `length` and `width` are built-in R functions, R Studio
might add "()" after length and width and if you leave the parentheses you will
get unexpected results. This is why you might see other programmers abbreviate
common words.

Create a third variable `area` and give it a value based on the current values
of `length` and `width`.

Show that changing the values of either `length` and `width` does not affect the
value of `area`.


### Functions and their arguments

Functions are "canned scripts" that automate more complicated sets of commands
including operations assignments, etc. Many functions are predefined, or can be
made available by importing R *packages* (more on that later). A function
usually gets one or more inputs called *arguments*. Functions often (but not
always) return a *value*. A typical example would be the function `sqrt()`. The
input (the argument) must be a number, and the return value (in fact, the
output) is the square root of that number. Executing a function ('running it')
is called *calling* the function. An example of a function call is:

```{r}
b <- sqrt(a)
```

Here, the value of `a` is given to the `sqrt()` function, the `sqrt()` function
calculates the square root, and returns the value which is then assigned to
the object `b`. This function is very simple, because it takes just one argument.

The return 'value' of a function need not be numerical (like that of `sqrt()`),
and it also does not need to be a single item: it can be a set of things, or
even a dataset. We'll see that when we read data files into R.

Arguments can be anything, not only numbers or filenames, but also other 
objects. Exactly what each argument means differs per function, and must be
looked up in the documentation (see below). Some functions take arguments which
may either be specified by the user, or, if left out, take on a *default* value:
these are called *options*. Options are typically used to alter the way the
function operates, such as whether it ignores 'bad values', or what symbol to
use in a plot.  However, if you want something specific, you can specify a value
of your choice which will be used instead of the default.

Let's try a function that can take multiple arguments: `round()`

```{r}
round(3.14159)
```

Here, we've called `round()` with just one argument, `3.14159`, and it has
returned the value `3`.  That's because the default is to round to the nearest
whole number. If we want more digits we can see how to do that by getting
information about the `round()` function.  We can use `args(round)` or look at
the help for this function using `?round`.

```{r}
args(round)
```

```{r}
?round
```

We see that if we want a different number of digits, we can
type `digits = 2` or however many we want.

```{r}
round(3.14159, digits = 2)
```

If you provide the arguments in the exact same order as they are defined you
don't have to name them:

```{r}
round(3.14159, 2)
```

And if you do name the arguments, you can switch their order:

```{r}
round(digits = 2, x = 3.14159)
```

But if you don't name the arguments, you can't switch their order:

```{r}
round(2, 3.14159)
```

It's good practice to put the non-optional arguments (like the number you're
rounding) first in your function call, and to specify the names of all optional
arguments.  If you don't, someone reading your code might have to look up the
definition of a function with unfamiliar arguments to understand what you're
doing.

### Your turn! 

### Exercise

Type in `?round` at the console and then look at the output in the Help pane.
What other functions exist that are similar to `round`?
How do you use the `digits` parameter in the round function?


### Vectors and data types

A vector is the most common and basic data type in R, and is pretty much
the workhorse of R. A vector is composed by a series of values, which can be
either numbers or characters. We can assign a series of values to a vector using
the `c()` function. For example we can create a vector of years in the 1990s and
assign it to a new object `years_90s`:

```{r}
years_90s <- c(1990, 1991, 1992, 1993, 1994, 1995, 1996, 1997, 1998, 1999)
years_90s  
```

A vector can also contain characters. For example, we can have a vector of 
country names:

```{r}
countries <- c("Beligum", "Denmark", "Germany", "France")
countries 
```

The quotes around "Belgium", etc. are essential here. Without the quotes R
will assume there are objects called `Beligum`, `Denmark`, `Germany`, and
`France`. As these objects don't exist in R's memory, there will be an error
message:

```{r}
countries <- c(Beligum, Denmark, Germany, France)
```


### Inspecting objects

There are many functions that allow you to inspect the content of a
vector. `length()` tells you how many elements are in a particular vector:

```{r}
length(years_90s)
length(countries)
```

An important feature of a vector, is that all of the elements are the same type
of data. The function `class()` indicate the class (the type of element) of an
object:

```{r}
class(years_90s)
class(countries)
```

The function `str()` provides an overview of the structure of an object and its
elements. It is a useful function when working with large and complex
objects:

```{r}
str(years_90s)
str(countries)
```

You can use the `c()` function to add other elements to your vector:

```{r}
possessions <- c("bicycle", "radio", "television")
possessions <- c(possessions, "mobile_phone") # add to the end of the vector
possessions <- c("car", possessions) # add to the beginning of the vector
possessions
```

In the first line, we take the original vector `possessions`,
add the value `"mobile_phone"` to the end of it, and save the result back into
`possessions`. Then we add the value `"car"` to the beginning, again saving the
result back into `possessions`.

We can do this over and over again to grow a vector, or assemble a dataset.
As we program, this may be useful to add results that we are collecting or
calculating.

An **atomic vector** is the simplest R **data type** and is a linear vector of a
single type. Above, we saw 2 of the 6 main **atomic vector** types  that R
uses: `"character"` and `"numeric"` (or `"double"`). These are the basic
building blocks that all R objects are built from. The other 4 **atomic vector**
types are:

* `"logical"` for `TRUE` and `FALSE` (the boolean data type)
* `"integer"` for integer numbers (e.g., `2L`, the `L` indicates to R that it's
an integer)
* `"complex"` to represent complex numbers with real and imaginary parts (e.g.,
`1 + 4i`) and that's all we're going to say about them
* `"raw"` for bitstreams that we won't discuss further

You can check the type of your vector using the `typeof()` function and
inputting your vector as the argument.

Vectors are one of the many **data structures** that R uses. Other important
ones are lists (`list`), matrices (`matrix`), data frames (`data.frame`),
factors (`factor`) and arrays (`array`).

### Your turn!

### Exercise 1

We’ve seen that atomic vectors can be of type character, numeric (or double),
integer, and logical. But what happens if we try to mix these types in a
single vector?

### Exercise 2

What will happen in each of these examples? (hint: use `class()`
to check the data type of your objects):

```{r}
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

Why do you think it happens?

### Exercise 3 

How many values in `combined_logical` are `"TRUE"` (as a character) in the
following example:

```{r}
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
combined_logical <- c(num_logical, char_logical)
```


You've probably noticed that objects of different types get converted into a
single, shared type within a vector. In R, we call converting objects from one
class into another class _coercion_. These conversions happen according to a hierarchy, whereby some types get preferentially coerced into other types.

__Can you draw a diagram that represents the hierarchy of how these data types are coerced?__


### Lists  
 
While the elements of vectors have to be of the same data type, a list is a
special vector in `R` that allows for you to store a variety of types of 
objects. If you have a vector, a matrix, a character, you can store all of them
into one list object!

The arguments to the list function are the components of the list. Where the 
components can be characters, vectors, matrices, or other data structures. 
Here, we create a list whose components are the three vectors we've been 
working with:

```{r}
my_first_list <- list(years_90s, countries)
my_first_list
```

We notice that when printing a list, the output looks a bit different. There 
are a whole bunch of brackets! Let's break them down. I like to think of a list
as a shelf with cubby holes. The cubby holes are the components of the list, 
but there are elements in each cubby.  

* To get to a specific component (cubby) of a list, you use the double brackets
next to the name of the list (e.g `my_first_list[[1]]`).  
* To access the elements inside each cubby, you then use single square brackets 
(e.g. `my_first_list[[1]][2]`). 


### Named Lists

```{r}
my_named_list <- list(title = "statistics", 
                      numbers = 1:10,
                      data = TRUE)
my_named_list
```

We see the output for a named list is slightly different. Instead of double 
brackets next to each component, there are now `$` and names of the variable. 
This will help you understand the structure of data.frames (coming up next)!

