---
title: "Starting with Data"
date: "`r Sys.Date()`"
output: html_document
---


# Importing data  

There are two ways to import data into R, (1) use the built-in GUI in the
Environment tab, or (2) write the code yourself. If you use option one, after
you make your selection, the GUI will output the code you *could have used* to 
import the dataset. 

Let's go with option two, since we are learning to write code! 

# Data import with the here package

Importing data requires for you to tell R how to get to the dataset you are 
interested in analyzing. This can be rather cumbersome, if you resort to telling
R exactly where the data lives on your computer (e.g.,
`C:\Users\atheo\Documents\Service\csu-math-ws\data`). The fortunate thing about
RStudio projects is that they **always** act as homebase. So, you only need to 
specify where the data live *relative* to where your RStudio project is 
stored. 

The specification of a path to get to the dataset can be a bit confusing, but
there is help! The **here** package provides an intuitive way to "find your
project's files, based on the current working directory" when the RStudio 
project is loaded. Specifically, the `here()` function in the here package 
accepts character inputs, specifying the path for the data. For example, if I 
had the `gapminder.csv` stored in the `data` folder of my RStudio project, the 
following code will navigate to the data: `here("data", "gapminder_data.csv")`
 
We are also going to use the `read_csv()` function from the **`readr`** package,
which is part of the **`tidyverse`**; learn more about the **`tidyverse`**
collection of packages [here](https://www.tidyverse.org/).
**`readr`** gets installed as part as the **`tidyverse`** installation.
When you load  the **`tidyverse`** (`library(tidyverse)`), the core packages
(the packages used in most data analyses) get loaded, including **`readr`**.

So, before we can use the `read_csv()` function, we need to load the package. 
We can use `here::here()` to use the `here()` function from the here package, 
without loading the package, so long as the here package has been installed! 

```{r}
library(tidyverse)

gapminder <- read_csv(
  here::here("data", "gapminder_data.csv")
)
```

If you were to type in the code above, it is likely that the `read.csv()`
function would appear in the automatically populated list of functions. This
function is different from the `read_csv()` function, as it is included in the
"base" packages that come pre-installed with R. Overall, `read.csv()` behaves
similar to `read_csv()`, with a few notable differences. First, `read.csv()`
coerces column names with spaces and/or special characters to different names
(e.g. `interview date` becomes `interview.date`). Second, `read.csv()`
stores data as a `data.frame`, where `read_csv()` stores data as a `tibble`.
We prefer tibbles because they have nice printing properties among other
desirable qualities. Read more about tibbles [here](https://tibble.tidyverse.org/).


### Inspecting data 

The second statement in the code above creates a data frame but doesn't output
any data because, as you might recall, assignments (`<-`) don't display
anything. (Note, however, that `read_csv` may show informational
text about the data frame that is created.) If we want to check that our data
has been loaded, we can see the contents of the data frame by typing its name:
`gapminder` in the console.


# Structure of data  

The data we will use is organized into data tables. When you imported the 
`gapminder_data.csv` data into R was saved as an object. You are able to inspect
the structure of these data object using functions built in to R 
(no packages necessary).  

Run the following code. What is output from each of the following commands?

```{r}
class(gapminder) ## What is the object class of the data?

dim(gapminder) ## What is the first number represent? What about the second number?

names(gapminder) ## What does this output refer to?

str(gapminder) ## What is the structure of the data?

summary(gapminder) ## What is the data type of each variable in our dataset?

typeof(gapminder) ## What is the storage mode of a data.frame?
```

When we inspect dataframes, or other objects in R, there are some general 
functions that are useful to check the content/structure of the data. 
Here are some:

- size:  
    + `dim(data)`: rows and columns  
    + `nrow(data)`: number of rows  
    + `ncol(data)`: number of columns  


- content:  
    + `head(data)`: first 6 rows  
    + `tail(data)`: last 6 rows  
    + `View(data)`: opens viewer window in separate tab  
    

- names:  
    + `colnames(data)`: column names of dataframe  
    + `rownames(data)`: row names of dataframe  


- summary of content:  
    + `str(data)`: structure of object and information about the columns  
    + `glimpse(data)`: similar information to `str`, but neater output  
    + `summary(data)`: summary statistics for each column  
  

*Note:* Many of these are generic functions that will work on other data types, 
not just on `data.frame`s. 


## Indexing and subsetting data frames

Our `gapminder` data frame has rows and columns (it has 2 dimensions).
In practice, we may not need the entire data frame; for instance, we may only
be interested in a subset of the observations (the rows) or a particular set
of variables (the columns). If we want to extract some specific data from it,
we need to specify the "coordinates" we want from it. Row numbers come first,
followed by column numbers.

```{r}
## first element in the first column of the tibble
gapminder[1, 1]

## first element in the 6th column of the tibble 
gapminder[1, 6]

## first column of the tibble (as a vector)
gapminder[[1]]

## first column of the tibble
gapminder[1]

## first three elements in the 6th column of the tibble
gapminder[1:3, 6]

## the 3rd row of the tibble
gapminder[3, ]

## equivalent to head_gapminder <- head(gapminder)
head_gapminder <- gapminder[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r}
# The whole tibble, except the first column
gapminder[, -1]          

# Equivalent to head(gapminder)
gapminder[-c(7:131), ]   
```

`tibble`s can be subset by calling indices (as shown previously), but also by
calling their column names directly:

```{r}
gapminder["village"]       # Result is a tibble

gapminder[, "village"]     # Result is a tibble

gapminder[["village"]]     # Result is a vector

gapminder$village          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct
names of the columns.


### Your turn! 

### Exercise 1

Create a tibble (`gapminder_100`) containing only the data in
row 100 of the `gapminder` dataset.

### Exercise 2 

Notice how `nrow()` gave you the number of rows in the tibble?  

  * Use that number to pull out just that last row in the tibble.
  * Compare that with what you see as the last row using `tail()` to make 
  sure it's meeting expectations.
  * Pull out that last row using `nrow()` instead of the row number.    
  * Create a new tibble (`gapminder_last`) from that last row.

### Exercise 3 

Using the number of rows in the gapminder dataset that you found in
Exercise 2, extract the row that is in the middle of the dataset. Store
the content of this middle row in an object named `gapminder_middle`.

### Exercise 4

Combine `nrow()` with the `-` notation above to reproduce the behavior of
`head(gapminder)`, keeping just the first through 6th rows of the gapminder
dataset. Compare the output with the `head()` function to confirm they are the 
same!


# Changing Data Type  

In most cases, it is preferable to leave variables encoded as text as 
characters, but there are instances where a different data type would be 
preferable. For example, what if you would like for the `continent`s to be 
ordered in a specific way? 

Consider the variables country and continent. These variables represent a
broader class of what we call categorical variables. In R there are two ways 
to store this information, (1) as a series of `character` strings, or (2) as a 
`factor`. In the early days of coding in R, factors were more efficient than 
characters, since you only need to store the level of the factor each 
observation went with. 

While factors are still useful in today's statistical analyses and data 
visualizations, they can be tricky to deal with. When you convert a variable 
to a factor, for many operations you will get different results than for a 
character (McNamara & Horton, 2017). 

In these data country has 142 levels and continent has five levels (Asia, 
Europe, Africa, Americas, Oceania). If we want R to view these variables
as factors instead of characters, we need to change their data type.   


```{r}  
## Tells you the unique values of country
unique(gapminder$country)

## Tells you the unique values of continent
unique(gapminder$continent)

## creates a new variable that is the factor version of country
gapminder$country_fct <- as.factor(gapminder$country) 

## creates a new variable that is the factor version of continent
gapminder$continent_fct <- as.factor(gapminder$continent) 

```

There is also a function that will allow for you to specify the order of the 
levels of a factor! As we saw before, the `as.factor()` function chooses the 
levels alphabetically. Suppose you would like for the continents to be in the 
following order: Americas, Europe, Asia, Africa, Oceania. 

Using the `factor()` function this would look like:

```{r}
gapminder$continent_fct <- factor(gapminder$continent, 
                                  levels = c("Americas", "Europe", "Africa", 
                                             "Oceania")) 
```

### Your turn! 

### Exercise 1 

Year was saved as an integer data type (1952 - 2007), but we may want to
consider it to be a categorical variable (a factor). Write the R code to create
a new variable called `year_fct` that is a factor of `year` (as you did with
`country` and `continent`). 

### Exercise 2 

Now, using the functions you learned for inspecting data, verify that `year_fct`
is viewed as a categorical variable, with the same levels as `year`. (hint: you
have already used three functions that would do this for you)  


### Cleaning Data  

In many instances, you will deal with data that are not "clean". Based on the
output below, how many rows in the `gapminder` data have an NA present?  

```{r NA, purl = TRUE, tidy = TRUE}
## gives the dimensions of the dataset in (row, column) format
dim(gapminder)   

## na.omit() takes dataframes, matrices, and vectors and returns the object with incomplete cases removed (NA's removed)
dim(na.omit(gapminder)) 
```


### Your turn!

### Exercise

Save a "clean" version of the `gapminder` dataset into an object named 
`gapminder_clean`. The "clean" version should contain no missing values. 

